**Does the Heart Lies: The Role of Interoception and Theory of Mind in Deception**

### 2022-11-14 paper/manuscript.Rmd
intro

---
title: '**The Heart Lies: The Role of Interoception and Theory of Mind in Deception**'
shorttitle        : "TITLE"
author:
  - name          : "Dominique Makowski"
    affiliation   : " 1, *"
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dmakowski@ntu.edu.sg"
  - name          : "Zen J. Lau"
    affiliation   : " 1"
  - name          : "Tam Pham"
    affiliation   : " 1"
  - name          : "S.H. Annabel Chen"
    affiliation   : " 1, 2, 3, 4"
    corresponding : yes    # Define only one corresponding author
    address       : "HSS 04-19, 48 Nanyang Avenue, Singapore"
    email         : "annabelchen@ntu.edu.sg"
affiliation:
  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "Lee Kong Chian School of Medicine, Nanyang Technological University, Singapore"
  - id            : "4"
    institution   : "National Institute of Education, Nanyang Technological University, Singapore"
abstract: |
output:
  papaja::apa6_word:
    keep_tex: yes
  pdf_document: default
  papaja::apa6_pdf:
    keep_tex: yes
keywords: ""
wordcount: ''
bibliography: ["references.bib"]
floatsintext: yes
figsintext: yes
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
documentclass: apa6
classoption: man
header-includes:
- \usepackage[labelfont=bf, font={color=gray,small}]{caption}
- \usepackage{float}
- \usepackage[document]{ragged2e}
csl: utils/apa.csl
---

@@

```{=html}
<!-- 
Journals:
Neuroscience Letters https://www.journals.elsevier.com/neuroscience-letters
Biological Psychology https://www.journals.elsevier.com/biological-psychology
Psychophysiology https://onlinelibrary.wiley.com/journal/14698986
Cognitive Neurodynamics https://www.springer.com/journal/11571
Neuroscience & Biobehavioral Reviews https://www.journals.elsevier.com/neuroscience-and-biobehavioral-reviews
-->
```
# Introduction

@@

In this post-truth era, deception has become increasingly prevalent in our everyday lives [@lewandowsky2017beyond]. Although the act of deceiving is regarded as complex and often requires effortful self-regulation to conceal [@abe2007deceiving; @krokoszinski2016emotion], research shows that people tend to under-estimate their ability to successfully lie [@elaad2003effects]. However, relative to the abundant literature regarding deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], surprisingly little attention has been conferred to investigating the inter-individual differences in people's lying ability. Given meta-analytic findings wherein the deceptive ability of the liar is suggested to be more important than the lie detector in deceptive interactions [@verigin2019lie; @bond2008individual; @levine2011sender], it is perhaps more pertinent to investigate what makes someone a more convincing liar.

@@

As deception requires the liar to intentionally manipulate the beliefs of others [@burgoon1994interpersonal; @sip2012if], most research on lying capability has been focused on higher order cognitive processes, particularly in theory of mind (TOM). TOM refers to the ability to infer that others have mental states, such as beliefs, emotions and intentions, distinct from ourselves [@lee2021lying; @wellman2001meta; @baron1997mindblindness]. Indeed, the capacity to tell lies, as well as the complexity of lies told, have previously been found to be related to higher TOM abilities [@talwar2007lying; @evans2011verbal; @talwar2017role]. Furthermore, the deceptive capabilities of children is often used as a measure of their skills in TOM. However, studies investigating the links between TOM and lying ability have predominantly been focused on children and individuals with neurocognitive deficits [@beaudoin2020systematic; @bora2017meta; @roheger2022progression]; likely a result of the view of TOM as the foundation of other higher order cognitive processes [@ketelaars2010dynamics; @isaksson2021s]. As such, little is known regarding the role TOM plays in the lying ability of healthy adults.

@@

Alternatively, interoceptive abilities could also influence inter-individual differences in lying. Broadly defined as one's sensitivity to their own internal signals and bodily states [@murphy2019classifying; @chen2021emerging; @weiss2014interaction], @garfinkel2015knowing's conceptualizes interoception as a three-dimensional framework comprising three distinct facets, namely - *interoceptive accuracy*, the objective ability to monitor internal bodily signals; *interoceptive sensibility*, the subjective confidence in one's interoceptive accuracy; and *interoceptive awareness*, one's ability to identify and appropriately respond to their perceived interoceptive states. Interoception has increasingly been tied to subjective perceptual experiences [@connell2018interoception; @seth2012interoceptive], as well as individual differences in executive functions, emotional processing and decision-making [@barrett2015interoceptive; @petzschner2021computational; @murphy2019classifying]. Although little research interest has been conferred to investigating the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. In fact, this is consistent with the somatic marker hypothesis, which posits interoception is associated with a propensity towards making rational decisions [@damasio1996somatic].

In contrast, some studies have instead found heightened interoceptive attention, one's self-focus towards internal bodily signals, to predict apathetic, immoral behaviour, such as cheating [@ditto2006visceral; @williams2016valjean; @lenggenhager2013listening]. Extending these findings to social cognition, @vabba2022interoceptive further reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high (relative to low), whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told. Taking these mixed findings to perspective with the paucity in research linking deceptive ability with interoception, more studies are warranted to comprehensively examine the relationship between interoception and deception.

@@

The aim of the present study was to explore the interplay effects of TOM and interoception on individuals' deceptive ability, as indicated by their lie confidence, physiological arousal and response time. To this end, we designed a directed-lying bi-modal feedback paradigm comprising 2 conditions ( *Interrogation* vs *Polygraph*) with distinct feedback cues. In particular, salient social feedback cues were conferred during the *Interrogation condition*, whereas visual biofeedback cues were given during the *Polygraph condition*. We expected interoception to positively predict lying ability in the *Polygraph condition*; participants with better interoception would exhibit higher lie confidence, shorter response time and lower physiological arousal. Alternatively, we hypothesized TOM would positively predict individuals' capability in lying in the *Interrogation condition*.

@@

Along with open science standards, the material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) for this research is available at [**https://github.com/DominiqueMakowski/DeceptionInteroTom**](https://github.com/DominiqueMakowski/DeceptionInteroTom){uri.}.

@@

>>>>

---
title: '**Does the Heart Lies: The Role of Interoception and Theory of Mind in Deception**'
shorttitle        : "TITLE"
author:

  - name          : "Dominique Makowski"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dom.makowski@gmail.com"
    orcid         : 0000-0001-5375-9967
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Data curation"
      - "Formal Analysis"
      - "Funding acquisition"
      - "Investigation"
      - "Methodology"
      - "Project administration"
      - "Resources"
      - "Software"
      - "Supervision"
      - "Validation"
      - "Visualization"
      - "Writing – original draft"
  - name          : "Zen J. Lau"
    affiliation   : " 1"
  - name          : "Tam Pham"
    affiliation   : " 1"
  - name          : "An Shu Te"
    affiliation   : "1"
    orcid         : 0000-0001-8008-2824
    role:
      - "Project administration"
      - "Resources"
      - "Investigation"
      - "Writing – original draft"
  - name          : "Stephanie Kirk"
    affiliation   : "1"
    orcid         : 0000-0002-9312-5552
    role:
      - "Project administration"
      - "Resources"
      - "Writing – original draft"
  - name          : "Ngoi Zi Liang"
    affiliation   : "1"
    role:
      - "Project administration"
      - "Resources"
      - "Writing – review & editing"
  - name          : "S.H. Annabel Chen"
    affiliation   : "1, 2, 3, 4"
    orcid         : 0000-0002-1540-5516
    role:
      - "Project administration"
      - "Supervision"
      - "Writing – review & editing"

affiliation:
  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "Lee Kong Chian School of Medicine, Nanyang Technological University, Singapore"
  - id            : "4"
    institution   : "National Institute of Education, Nanyang Technological University, Singapore"
abstract: |
output:
  papaja::apa6_word:
    keep_tex: yes
  pdf_document: default
  papaja::apa6_pdf:
    keep_tex: yes
keywords: ""
wordcount: ''
bibliography: ["references.bib"]
floatsintext: yes
figsintext: yes
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
documentclass: apa6
classoption: man
header-includes:
- \usepackage[labelfont=bf, font={color=gray,small}]{caption}
- \usepackage{float}
- \usepackage[document]{ragged2e}
csl: utils/apa.csl
---

@@

<!-- 
Journals:
Neuroscience Letters https://www.journals.elsevier.com/neuroscience-letters
Biological Psychology https://www.journals.elsevier.com/biological-psychology
Psychophysiology https://onlinelibrary.wiley.com/journal/14698986
Cognitive Neurodynamics https://www.springer.com/journal/11571
Neuroscience & Biobehavioral Reviews https://www.journals.elsevier.com/neuroscience-and-biobehavioral-reviews
-->

@@

Lying - the intentional attempt at ... [**Insert definition**] - is a prevalent phenomenon carrying potentially important consequences. Interestingly, evidence suggests that the successful detection of a lying attempt depends more on the ability of the liar, than on the performance of the lie detector [@verigin2019lie; @bond2008individual; @levine2011sender]. However, the factors contributing to one's deception abilities remain unclear, most of the deception literature focusing instead on deception detection [@viji2022history; @sternglanz2019review; @masip2017deception].

<!-- research shows that people tend to under-estimate their ability to successfully lie [@elaad2003effects] -->

@@

As deception requires the liar to intentionally manipulate the beliefs of others [@burgoon1994interpersonal; @sip2012if], a significant line of research has been focused on the role of theory of mind (ToM) in lying ability. ToM refers to the ability to infer that others have mental states, such as beliefs, emotions and intentions, distinct from ourselves [@lee2021lying; @wellman2001meta; @baron1997mindblindness]. The ability to tell lies, as well as their complexity, have previously been found to be related to higher TOM abilities [@talwar2007lying; @evans2011verbal; @talwar2017role]. However, studies investigating the link between ToM and deception have predominantly been focused on children and neuroatypical individuals [@beaudoin2020systematic; @bora2017meta; @roheger2022progression], and its importance in healthy adults remains to be clarified.

@@

Besides paying to attention to the person we lie to, trying to gauge whether they believe us, it is also possible that some attention is directed inwards: monitoring our own body and its reactions (e.g., cardiac activity and its related changes such as blushing), susceptible to be used as cues to infer our real intent. This begs the question of the potential role of interoceptive abilities in deception ability. Broadly defined as one's sensitivity to their own internal signals and bodily states [@murphy2019classifying; @chen2021emerging; @weiss2014interaction], @garfinkel2015knowing's conceptualizes interoception as a three-dimensional framework comprising three distinct facets, namely, *interoceptive accuracy* - the objective ability to monitor internal bodily signals; *interoceptive sensibility* - the subjective confidence in one's interoceptive accuracy; and *interoceptive awareness* - one's ability to identify and appropriately respond to their perceived interoceptive states. Interoception has increasingly been tied to subjective perceptual experiences [@connell2018interoception; @seth2012interoceptive], as well as individual differences in executive functions, emotional processing and decision-making [@barrett2015interoceptive; @petzschner2021computational; @murphy2019classifying]. Although little research interest has been conferred to investigating the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. In fact, this is consistent with the somatic marker hypothesis, which posits an association between interoception and a propensity towards making rational decisions [@damasio1996somatic].

In contrast, some studies have instead found heightened interoceptive attention, one's self-focus towards internal bodily signals, to predict apathetic, immoral behaviour, such as cheating [@ditto2006visceral; @williams2016valjean; @lenggenhager2013listening]. Extending these findings to social cognition, @vabba2022interoceptive further reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high, whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told. In a context of scarce research on interoception and deception, more studies are needed to clarify these these mixed findings.

@@

The aim of the present study was to explore the contribution of ToM and interoception abilities on individuals' deceptive skills, as indicated by their lying confidence, physiological arousal and response time. To this end, we designed a directed-lying paradigm with 2 conditions differing in the nature of their feedback cues. The *Interrogation* condition was designed to emphasize (and preferentially mobilize) ToM-related mechanims, whereas the *Polygraph* condition was designed to emphasize interoceptive mechanisms. We expected ToM and interoception to positively predict lying ability (i.e., higher lie confidence, shorter response time and lower physiological arousal), in particular in the *Polygraph* and the *Interrogation* condition, respectively.

@@

@@

# Data Availability

The material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) for this research is available at [**https://github.com/DominiqueMakowski/DeceptionInteroTom**](https://github.com/DominiqueMakowski/DeceptionInteroTom){uri.}.
